Regression Cube Analysis of Cohort Study Data

### 2015-03-30 draft.tex
added breast fat evaluation

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al. \cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates and time plots to display and select the data.
%%
%Medical image data can be incorporated to investigate potentially interesting individuals.
%%
%%We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.
%%
Generalized Pairs plots (\texttt{GPLOM'S}) extend the idea of scatter plot matrices by the pairwise depiction of heterogeneous data using type-combination-dependent visualizations \cite{GPLOMS, Francois2013}.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter and are therefore suitable for \emph{small} epidemiological data sets.
%%
Dai et al. \cite{Dai2005} incorporate a \texttt{GPLOM}-like visualization using choropleth maps mapping spatial data, such as mortality rates together with scatter plots augmented with \emph{Pearsson's} $r$ values.
%%
A \emph{concept map} summarizes features related to a specified disease.
%%
Time-dependent epidemiological data is visualized by Chui et al. \cite{Chui2011} using multi-panel graphs highlighting risk factor differences with age and gender towards influenza- and salmonellosis-associated hospitalizations.
%%
%We also include confounder features (age, gender) into the analysis while focusing on non-time-dependent cohort study data sets.
%%
\paragraph{Statistical Analysis.}
%%
%Statistical analysis is an inherent part of Visual Analysis.
%%
%Here we present work showing a strong focus on statistical analysis in this context.
%% Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization
Bertini et al. \cite{Bertini} present an overview over quality metrics describing high dimensional data.
%%
Their research agenda acts as guideline for our design, comprising perceptual tuning towards human pattern recognition of important aspects, scalability between different data set sizes and application evaluation with domain experts.
%%
%% TODO: Use this one? "Statistical resilience of random populations to random perturbations" - http://www.ncbi.nlm.nih.gov/pubmed/19256997 - wie eine Variable sich √É¬§ndert, wenn man per Zufall die Populationen ver√É¬§ndert, um zu schauen wie zuverl√É¬§ssig diese Statistik ist.
%% The Sparse Regression Cube: A Reliable Modeling Technique for Open Cyber-physical Systems
Ahmadi et al. \cite{Ahmadi} define the \emph{Sparse Regression Cube}, which partitions sparse high-dimensional data into subspaces, which are then described by their most reliable linear regression model.
%%
%In their application, they try to find most fuel-efficient roads connecting two user-defined landmarks.
%%
They focus on an algebraic representation for efficient regression-model calculation to find the best fit for a subspace.
%%
%The dimensions for the calculations are fixed, while we assess different features for each model calculation and can therefore not benefit from a hierarchical calculation approach.

@@

%Hielscher et al. \cite{Hielscher} assess subject similarity towards hepatic steatosis (fatty liver disease) to identify common risk factors using [...]?
%\com{Uli, can you please complete this, I do not have the pdf available!}
%% Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis
Niemann et al. \cite{Niemann2014} investigate risk factors of hepatic steatosis using decision trees with interactive data mining tools.
%%
They extracted association rules that serve as basis for a proof-of-concept evaluation.
%%
In \cite{Niemann2015}, Niemann et al. improve on classification performance by generating features (called \emph{evolution features}) that describe latent temporal information across the study waves.
%%
We try to reproduce their results and investigate their findings further.
%%
%\com{Gla√É¬üer, S., Niemann, U., Preim, B., and Spiliopoulou, M. (2013). Can we Distinguish Between Benign and Ma- lignant Breast Tumors in DCE-MRI}

@@

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The ability of a system to discover knowledge is difficult to measure.
%%
Lam et al. \cite{Lam2012} propose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out using case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%
% folgend ist doch korrekt oder?
We employ \emph{VDAR} for the evaluation of our system.
%%
\paragraph{Participants, Setup and Procedure.}
%%
%One major advantage of focusing on web-based research was the possibility to bridge the geographical gap towards the epidemiological domain experts.
%%
We conducted a web-based analysis by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.

@@

%%
% TODO du stellst aber 3 vor...
We conducted a study with two participants, who also co-authored this publication.
%%
Domain expert for the breast fat data set analysis is \emph{KH}, a clinician (10 years of experience) with focus on epidemiological research.
%%
\emph{KH} is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.

@@

% OK so?
The third participant was 
\emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used in \cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
% ok?  to indicate absence or presence of the disease.
Liver fat concentrations of no more than 10\% are mapped to the 'negative' class; values greater than 10\% are mapped to the 'positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants for which the MRI data was available to that point.
%%
The MRI scans for each subject have been introduced in SHIP-2, so the target feature is only present for one of three study waves.

@@

The data comprises 1,186 female subjects (368 from \texttt{SHIP-2}, 818 from \texttt{SHIP-TREND-0} cohort).
%%
It contains 231 dimensions, holding information about somatometric features (e.g. body size or weight) consumption behavior (e.g. alcohol or tobacco), personal and medical history (e.g., occupation or prior diseases), women specific features (e.g., number of born children or contraception type) as well as mammography features.
% It contains 231 dimensions, holding information about
% \begin{itemize}
% 	\item \emph{somatometric features}, e.g., body size, weight or BMI,
% 	\item \emph{lifestyle features}, e.g., alcohol/tobacco consume,
% 	\item \emph{personal history}, e.g., occupation, marital status,
% 	\item \emph{medical history}, e.g., current or prior diseases or laboratory values, such as cholesterol levels,
% 	\item \emph{women specific features}, e.g., number of born children, contraception type or hormone replacement therapy and
% 	\item \emph{mammography features} derived from the image data.
% \end{itemize}
%%
The latter were derived from manual segmented MRI data for each subject carried out by radiologists.
%%
%%
%\com{Describe Parenchym tissue and its relation to breast cancer.}
%\com{Describe how the data was acquired and preprocessed}

@@

% nicht mit 3.ZF anfangen, das klingt niht gut.
The Zoom, Filter and Analyze Further Step (\textbf{3.ZF}) was realised by slicing through the cube using the mouse input to inspect the hot spots. Anaylzing the mosaic plot (\textbf{4.DD}) revealed high correlations for somatometric features, hepatic steatosis indicator-features as well as laboratory values, such as creatinine (used renal retention parameter) and uric acid (among others used as gout and diabetes risk factors) magnitudes.

@@

% da fehlte was im satz

@@

\emph{UN} could reproduce most results, some featuresexhibit lower correlations though, e.g., creatinine magnitudes.

@@

@@

% das ist doch super. Hier noch ein Satz ob die Radiologen das gut finden. ist das nicht richtig geil? Mehr Infos!!!

@@

\textbf{Interleukin-6 correlation with liver fat.}
%%
During the analysis, one hot spot was always observable in the \emph{SI} and \emph{ZF} incorporating a high \emph{Interleukin-6} (\emph{IL-6}) correlation with liver fat values (Fig.~\ref{fig:Application} b).
%%
The correlation was high for both the dichotomized and continuous target feature.
%%
% hier fehlt was
In literature, relations between liver cancer \cite{He2013} and also chronic liver diseases \cite{Streetz2003}.
%%
For mice, strong effects of \emph{IL-6} with hepatic steatosis were described \cite{Hong2004}.
%%
The finding is subject of further analysis.
%%
\subsection{Case 2: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
The analysis aims to find relationships on the breast cancer data using features derived from the mammography analysis.
%%
Relationships between the share of parenchyma tissue on the overall breast volume are of high interest \cite{Mccormack2006}.
%%
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
Our method was well received among the domain experts.
%%
For the first time, they were able to derive an overview visualization custom tailored to underlying assumptions.
%%
In this context, \emph{TI} highlighted the ability to simultaneously analyze thousands of regression models while maintaining little time expenses for rating them.
%The novel way of modeling knowledge to derive hypothesis driven overview visualizations allow for detailed analysis of target features, either disease indicators or even image-derived features.
%%
%TI commented on the

>>>>

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al. \cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates and time plots to display and select the data.
%%
%Medical image data can be incorporated to investigate potentially interesting individuals.
%%
%%We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.
%%
Generalized Pairs plots (\texttt{GPLOM'S}) extend the idea of scatter plot matrices by the pairwise depiction of heterogeneous data using type-combination-dependent visualizations \cite{GPLOMS, Francois2013}.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter and are therefore suitable for \emph{small} epidemiological data sets.
%%
Dai et al. \cite{Dai2005} incorporate a \texttt{GPLOM}-like visualization using choropleth maps mapping spatial data, such as mortality rates together with scatter plots augmented with \emph{Pearsson's} $r$ values.
%%
A \emph{concept map} summarizes features related to a specified disease.
%%
Time-dependent epidemiological data is visualized by Chui et al. \cite{Chui2011} using multi-panel graphs highlighting risk factor differences with age and gender towards influenza- and salmonellosis-associated hospitalizations.
%%
%We also include confounder features (age, gender) into the analysis while focusing on non-time-dependent cohort study data sets.
%%
\paragraph{Statistical Analysis.}
%%
%Statistical analysis is an inherent part of Visual Analysis.
%%
%Here we present work showing a strong focus on statistical analysis in this context.
%% Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization
Bertini et al. \cite{Bertini} present an overview over quality metrics describing high dimensional data.
%%
Their research agenda acts as guideline for our design, comprising perceptual tuning towards human pattern recognition of important aspects, scalability between different data set sizes and application testing with domain experts.
%%
%% TODO: Use this one? "Statistical resilience of random populations to random perturbations" - http://www.ncbi.nlm.nih.gov/pubmed/19256997 - wie eine Variable sich √É¬§ndert, wenn man per Zufall die Populationen ver√É¬§ndert, um zu schauen wie zuverl√É¬§ssig diese Statistik ist.
%% The Sparse Regression Cube: A Reliable Modeling Technique for Open Cyber-physical Systems
Ahmadi et al. \cite{Ahmadi} define the \emph{Sparse Regression Cube}, which partitions sparse high-dimensional data into subspaces, which are then described by their most reliable linear regression model.
%%
%In their application, they try to find most fuel-efficient roads connecting two user-defined landmarks.
%%
They focus on an algebraic representation for efficient regression-model calculation to find the best fit for a subspace.
%%
%The dimensions for the calculations are fixed, while we assess different features for each model calculation and can therefore not benefit from a hierarchical calculation approach.

@@

%Hielscher et al. \cite{Hielscher} assess subject similarity towards hepatic steatosis (fatty liver disease) to identify common risk factors using [...]?
%\com{Uli, can you please complete this, I do not have the pdf available!}
%% Learning and inspecting classification rules from longitudinal epidemiological data to identify predictive features on hepatic steatosis
Niemann et al. \cite{Niemann2014} investigate risk factors of hepatic steatosis using decision trees with interactive data mining tools.
%%
They extracted association rules that serve as basis for our proof-of-concept tests.
%%
In \cite{Niemann2015}, Niemann et al. improve on classification performance by generating features (called \emph{evolution features}) that describe latent temporal information across the study waves.
%%
We try to reproduce their results and investigate their findings further.
%%
%\com{Gla√É¬üer, S., Niemann, U., Preim, B., and Spiliopoulou, M. (2013). Can we Distinguish Between Benign and Ma- lignant Breast Tumors in DCE-MRI}

@@

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The ability of a system to discover knowledge is difficult to measure.
%%
Lam et al. \cite{Lam2012} propose the \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique, which is focused on the characterization of a systems' ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out using case studies using thinking-aloud techniques to comprehend the reasoning and thought process of the user.
%
% folgend ist doch korrekt oder?
We employ the \emph{VDAR} technique for analyzing our system.
%%
\paragraph{Participants, Setup and Procedure.}
%%
%One major advantage of focusing on web-based research was the possibility to bridge the geographical gap towards the epidemiological domain experts.
%%
We conducted a web-based analysis by using an online-meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.

@@

%%
We conducted a three with two participants. %, who also co-authored this publication.
%%
Domain expert for the breast fat data set analysis is \emph{KH}, a clinician (10 years of experience) with focus on epidemiological research.
%%
\emph{KH} is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.

@@

% OK so?
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used in \cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
% ok?  to indicate absence or presence of the disease.
Liver fat concentrations of no more than 10\% are mapped to the 'negative' class; values greater than 10\% are mapped to the 'positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants for which the MRI data was available to that point.
%%
The MRI scans for each subject have been introduced in SHIP-2, so the target feature is only present for one of three study waves.

@@

The data comprises 1,186 female subjects (368 from \texttt{SHIP-2}, 818 from \texttt{SHIP-TREND-0} cohort).
%%
It contains 231 dimensions, holding information about somatometric features (e.g. body size or weight) consumption behavior (e.g. alcohol or tobacco), personal and medical history (e.g., occupation or prior diseases), women specific features (e.g., number of born children or contraception type) as well as mammography features (e.g., fat content or parenchyma tissue proportion to volume).
% It contains 231 dimensions, holding information about
% \begin{itemize}
% 	\item \emph{somatometric features}, e.g., body size, weight or BMI,
% 	\item \emph{lifestyle features}, e.g., alcohol/tobacco consume,
% 	\item \emph{personal history}, e.g., occupation, marital status,
% 	\item \emph{medical history}, e.g., current or prior diseases or laboratory values, such as cholesterol levels,
% 	\item \emph{women specific features}, e.g., number of born children, contraception type or hormone replacement therapy and
% 	\item \emph{mammography features} derived from the image data.
% \end{itemize}
%%
The latter were derived from manual segmented MRI data for each subject carried out by radiologists.
%%
%%
%\com{Describe Parenchym tissue and its relation to breast cancer.}
%\com{Describe how the data was acquired and preprocessed}

@@

The Zoom, Filter and Analyze Further Step (\textbf{3.ZF}) was realized by slicing through the cube using the mouse input to inspect the hot spots.

@@

Analyzing the mosaic plot (\textbf{4.DD}) revealed high correlations for somatometric features, hepatic steatosis indicator-features as well as laboratory values, such as creatinine (used renal retention parameter) and uric acid (among others used as gout and diabetes risk factors) magnitudes.

@@

\emph{UN} could reproduce most results, some features exhibit lower correlations though, e.g., creatinine magnitudes.

@@

%% TODO Ergebnisse nachfragen: das ist doch super. Hier noch ein Satz ob die Radiologen das gut finden. ist das nicht richtig geil? Mehr Infos!!!

@@

@@

\textbf{Interleukin-6 correlation with liver fat.}
%%
During the analysis, one hot spot was always observable in the \emph{SI} and \emph{ZF} incorporating a high \emph{Interleukin-6} (\emph{IL-6}) correlation with liver fat values (Fig.~\ref{fig:Application} b).
%%
The correlation was high for both the dichotomized and continuous target feature.
%%
The literature described relations between \emph{IL-6} and liver cancer \cite{He2013} as well as chronic liver diseases \cite{Streetz2003}.
%%
For mice, strong effects of \emph{IL-6} with hepatic steatosis were described \cite{Hong2004}.
%%
The finding is subject of further analysis.
%%
\subsection{Case 2: Hypothesis-free Analysis of the Breast Cancer Data Set}
%%
The analysis aims to find relationships on the breast fat data using features derived from the mammography analysis.
%%
Relationships between the share of parenchyma tissue on the overall breast volume are of high interest \cite{Mccormack2006}.
%%

The \textbf{1.AF} was started by \emph{KH} using the default formula for hypothesis-free analysis ($Z \sim X + Y$).
%%
At first, she was interested in correlations with the parenchyma tissue segmentations, which was selected through the drop-down for the z-axis (\textbf{2.SI}).
%%
She observed expected strong correlations with age, body fat percentage, hip and waist circumference
as well as menstrual period or pregnancy status.
%%
Fat women also have a larger breast fat percentage, which also correlates with other somatometric features.
%%
Age is a strong influencing factor, as breast tissue degrades over time and subsequently the parenchyma tissue.
%%
\emph{KH} proceeded using [\textbf{2.SI}] and [\textbf{3.DD}] to check for relationships for different target features, such as current hormone replacement therapy, BI-RADS (classification of the mammography findings) as well as different diseases, such as diabetes or gout.
%%
She observed relationships matching her expectations and expert knowledge.
%%
One unexpected relationship was observed between mammography lesions and menstruation cycle towards spiral contraception were observed ($R^2$ of $0.77$).
%%
\emph{KH} proceeded with a detailed analysis of the parenchyma tissue.
%\emph{KH} derived at the hypotheses: \emph{Somatometric features impose the major influences on the parenchyma tissue.}
%%

\textbf{Detailed Breast Parenchyma Analysis.}
%%
The analysis was conducted by calculating the cube $Parenchyma-Percentage \sim X + Y + Z$.
%%
Using the 3D cube, \emph{KH} observed several hotspots.
%%
Navigating to them using the slicing facility of the 3D visualization highlighted several features of high influence, such as image-derived features, as glandular tissue density, parenchyma segmentation metrics.
%%
Also, strong correlations were observed in the diabetes slice, confirming expectations of \emph{KH} w.r.t. to its strong influence to the parenchyma tissue.
%%
A surprising find were the strong correlations with kidney disorder ($R^2$ values around $0.9$).
%%
The \textbf{3.DD} analysis however showed only 8 subjects with kidney disorder.
%%
The finding was noted and investigated further in the near future using a extensive data set.
%%
Lastly \emph{KH} assessed the influence of contraception related features, such as use birth control pills or the spiral, but found no significant correlations with the parenchyma.
%%
Features, such as alcohol intake yield no elevated $R^2$ values.
%%
\emph{KH} remarked that these features are believed to have an impact on the parenchyma tissue, but they are less reliability since they are self-reported.
%%
%mamma anamnesis breast fed, contraception (spirale, pille), alcohol intake
%Lastly, \emph{KH} assessed the influence of image-derived features on the parenchyma tissue.
%%
%She observed expected results 
%Using the slicing mode to fast 
%Hypotheses
%Detailed analysis: Cholesterin, influence of image-derived feautres, classification of the BI-Rads (how well did the manual segmentation perform)
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
Our method was well received among the domain experts.
%%
For the first time, they were able to derive an overview visualization custom tailored to underlying assumptions.
%%
In this context, \emph{TI} highlighted the ability to simultaneously analyze thousands of regression models while maintaining little time expenses for rating them.
%The novel way of modeling knowledge to derive hypothesis driven overview visualizations allow for detailed analysis of target features, either disease indicators or even image-derived features.
%%
%TI commented on the
